---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# terrainSliceR

<!-- badges: start -->
[![R CMD Check](https://github.com/mikejohnson51/terrain_sliceR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mikejohnson51/terrain_sliceR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of terrainSliceR is to generate DEM-based cross sections for hydrograhic networks.

## Installation

You can install the development version of terrainSliceR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mikejohnson51/terrain_sliceR")
```

# Example

This is a basic example which shows you how to cut cut cross sections for a network.

## Define Network


```{r example}
library(terrainSliceR)
library(dplyr)

(net = linestring %>% 
  mutate(bf_width = exp(0.700	 + 0.365* log(totdasqkm))))


plot(net$geometry)
```

## Cut cross sections

```{r}
(transects = cut_cross_sections(net = net,
                         id = "comid", 
                         bf_widths = pmax(50, net$bf_width * 7),
                         num = 10) )

plot(transects$geometry)
```

## Define Cross section points

```{r}
(pts = cross_section_pts(transects, 
                        dem = '/Volumes/Transcend/ngen/DEM-products/dem.vrt'))
```

## Classify Cross section points

```{r}
(classified_pts = classify_points(pts))
```

## Explore!

```{r}
library(ggplot2)

ggplot(data = filter(classified_pts, hy_id == 101) ) + 
  geom_point(aes(x = relative_distance, y = Z, color = class)) + 
  facet_wrap(~cs_id, scales = "free") + 
  theme_minimal() + 
  theme(legend.position = "bottom")
```

