---
title: "Basic Use"
description: |
  "Adding Cross Sections to Hydrofabric Datasets"
author:
  - name: "Mike Johnson"
    url: https://github.com/mikejohnson51
    affiliation: Lynker, NOAA-Affiliate
    affiliation_url: https://lynker.com
output: distill::distill_article
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

```{r setup}
library(hydrofabric3D)
```

The goal of hydrofabric3D is to generate DEM-based cross sections for hydrographic networks.

## Installation

You can install the development version of hydrofabric3D from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("mikejohnson51/hydrofabric3D")
```

# Example

This is a basic example which shows you how to cut cross sections for a network.

## Define Network

```{r example}
library(hydrofabric3D)
library(dplyr)

(net = linestring %>% 
  mutate(bf_width = exp(0.700	 + 0.365* log(totdasqkm))))


plot(net$geometry)
```

## Cut cross sections

```{r}
(transects = cut_cross_sections(net = net,
                         id = "comid", 
                         cs_widths = pmax(50, net$bf_width * 7),
                         num = 10) )

plot(transects$geometry)
```

## Define Cross section points

```{r}
(pts = cross_section_pts(transects, 
                        dem = "/vsicurl/https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/1/TIFF/USGS_Seamless_DEM_1.vrt"))
```

## Classify Cross section points

```{r}
(classified_pts = classify_points(pts))
```

## Explore!

```{r}
library(ggplot2)

ggplot(data = filter(classified_pts, hy_id == 101) ) + 
  geom_line(aes(x = relative_distance, y = Z)) + 
  geom_point(aes(x = relative_distance, y = Z, color = class)) + 
  facet_wrap(~cs_id, scales = "free") + 
  theme_minimal() + 
  theme(legend.position = "bottom")
```

## Time to get `r nrow(transects)` transects and `r nrow(classified_pts)` classified points ...

```{r}
system.time({
  cs = net %>% 
  cut_cross_sections(id = "comid", 
                     cs_widths = pmax(50, net$bf_width * 7),
                     num = 10) %>% 
  cross_section_pts(dem = '/vsicurl/https://prd-tnm.s3.amazonaws.com/StagedProducts/Elevation/1/TIFF/USGS_Seamless_DEM_1.vrt') %>% 
  classify_points()
})
```

